  ---
  title: "World Happiness Report from 2005-2024 By R Markdown"
  author: "Moath"
  date: "2024-07-02"
  output: html_document
  ---
  // ignore_for_file: prefer_const_constructors, prefer_interpolation_to_compose_strings, avoid_print

  ```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  ```

  # installing and loading the required packages  
  ```{r, Loading the Dataset}
  install.packages("tidyverse")
  install.packages("ggplot2")
  install.packages("dplyr")
  install.packages("lsa")

  library(tidyverse)
  library(ggplot2)   
  library(dplyr)      
  library(skimr)     
  library(kableExtra)
  library(lsa)
  library(corrplot)
  ```



  # Read the Dataset  
  ```{r, Reading the Dataset}
  library(readr)

  World_happiness_data_2024 <- read.csv("path/to/your/dataset/World-happiness-report-2024.csv")
  World_happiness_data <- read.csv("path/to/your/dataset/World-happiness-report.csv")


  str(World_happiness_data_2024)

  #.......................

  str(World_happiness_data)
  ```




  # let's take a brief look of the "World_happiness_data"
  ```{r, brief of the dataset}

  head(World_happiness_data)
  summary(World_happiness_data)
  ```

  # let's take a brief look of the "World_happiness_data_2024"
  ```{r, brief of the dataset}

  head(World_happiness_data_2024)
  summary(World_happiness_data_2024)
  ```


  # Here is a full view of the dataset after sorting it by the "Life.Ladder" in an Ascending order.
  ```{r, full view of the dataset}

  arranged_data <- World_happiness_data %>%  arrange(Life.Ladder)

    head(arranged_data)

  ```



  ## After further investigtion, I have found some NULL values in the datasete.I will summarize the columns that contains atleast one row with a NULL value:  
  ```{r}

  null_count <- World_happiness_data %>% summarise_all(~sum(is.na(.)))

  null_columns_names <- null_count %>% select_if(~.>0)

  print(null_columns_names)
  ```


  # The same for World_happiness_data_2024
  ```{r}

  null_count_2024 <- World_happiness_data_2024 %>% summarise_all(~sum(is.na(.)))

  null_columns_names_2024 <- null_count_2024 %>% select_if(~.>0)

  print(null_columns_names_2024)

  ```



  # I have chosen to replace the NULL values with the "Mean/Average Imputation" because This can be suitable for numerical variables like (GDP per capita, life expectancy, or other continuous measures). It's straightforward and preserves the overall distribution of the data.

  # More Precisely: I will calculate the Mean/Average for the values of each column for each Country Group alone across the years, then replace the NULL values with it. 

  # For Example: "Afghanistan" had two NULL values in the "Log.GDP.per.capita" column, I have Replaced these null values with the "Mean/Average" of values in the "Log.GDP.per.capita" for "Afghanistan" country Only across the years.

#In case one country had all the values as NULL, I will replace them with the "Mean/Average" of the whole column.


  ```{r}
  library(dplyr)

  mean_excluding_na_nan <- function(x) {
    valid_values <- x[!is.na(x) & !is.nan(x)]
    mean(valid_values)
  }

  # Group by Country.name and fill NA and NaN values with the mean of each country
  world_happiness_clean <- World_happiness_data %>%
    group_by(Country.name) %>%
    mutate(across(where(is.numeric), 
                  ~ifelse(is.na(.) | is.nan(.), mean_excluding_na_nan(.), .)))

  # Repeat the process without grouping for the countries that has all values in it as NA

  world_happiness_clean_final <- world_happiness_clean %>%
    ungroup() %>%
    mutate(across(where(is.numeric), 
                  ~ifelse(is.na(.) | is.nan(.), mean_excluding_na_nan(.), .)))

  # My Analysis without will be without the country of "Israel"

  world_happiness_clean_final <- world_happiness_clean_final %>%
    filter(Country.name != "Israel")



  ```




  # I am gonna repeat the process for the "World_happiness_data_2024" without the grouping, but with the "Mean/Average" for the whole column. because this dataset contains only one row of data for each country.

  ```{r}
  library(dplyr)
  library(lsa)


  mean_excluding_na_nan <- function(x) {
    valid_values <- x[!is.na(x) & !is.nan(x)]
    
    mean(valid_values)
  }


  world_happiness_clean_2024 <- World_happiness_data_2024 %>%
    mutate(across(where(is.numeric),
    .fns = ~ifelse(is.na(.) | is.nan(.) | . ==0, mean_excluding_na_nan(.), .)))

    world_happiness_clean_2024 <- world_happiness_clean_2024 %>%
    filter(Country.name != "Israel")

  head(world_happiness_clean_2024)

  ```



  # Let's draw the number of regions per country
  ```{r}
  install.packages("plotly")
  library(ggplot2)
  library(plotly)

  p <- ggplot(world_happiness_clean_2024, aes(x = Regional.indicator, fill = Regional.indicator)) + 
    geom_bar() +
    labs(title = "Number of Countries per Region",
        x = "Region",
        y = "Number of Countries") +
    theme_minimal() +
    theme(legend.title = element_blank(),
          axis.text.x = element_text(angle = 90, hjust = 1)) 

  interactive_plot <- ggplotly(p)

  interactive_plot

  ```



  # Let's see the happiness score across the different regions
  ```{r}
  library(RColorBrewer)


  colors <- brewer.pal(n = length(unique(world_happiness_clean_2024$Regional.indicator)), name = "Set3") 

  ggplot(world_happiness_clean_2024, aes(x = Regional.indicator, y = Ladder.score, fill = Regional.indicator)) + 
    geom_bar(stat = "identity") +
    labs(title = "Happiness (Ladder score) Across Regions",
        x = "Region",
        y = "Happiness (Ladder score)") +
    scale_fill_manual(values = colors) +  
    theme_minimal(base_size = 14) +  
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 12),  
          axis.text.y = element_text(size = 12), 
          axis.title = element_text(size = 14),  
          legend.position = "none")

  ```




  #Now let's get beck to "world_happiness_clean" and see the correlation between all the factors by Drawing a Correlation Matrix (Heatmap)
  ```{r}

  numeric_columns <- select_if(world_happiness_clean_main, is.numeric) %>% select(-year) 


  correlation_matrix <- cor(numeric_columns, use = "complete.obs")

  upper_triangle_mask <- matrix(NA, nrow = nrow(correlation_matrix), ncol = ncol(correlation_matrix))
  upper_triangle_mask[upper.tri(upper_triangle_mask)] <- TRUE

  corrplot(correlation_matrix, method = "color", 
          col = colorRampPalette(c("blue", "white", "Brown"))(200),
          type = "lower", 
          tl.col = "black", 
          tl.srt = 45, 
          addCoef.col = "black", 
          number.cex = 0.7, 
          diag = FALSE)

  ```




  # Now let's see what are the most correlated factors with the Life Ladder Score
  ```{r}
  numeric_columns <- select_if(world_happiness_clean_main, is.numeric) %>% select(-year) 

  # Calculate correlations with Life Ladder
  correlations <- cor(numeric_columns, use = "complete.obs")
  cor_life_ladder <- correlations["Life.Ladder", -1]  

  # Prepare data for visualization
  cor_df <- data.frame(
    factor = names(cor_life_ladder),
    correlation = as.numeric(cor_life_ladder)
  )

  # Create bar plot to visualize the correlations
  p <- ggplot(cor_df, aes(x = reorder(factor, correlation), y = correlation, fill = correlation)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = "Factors Influencing Life Ladder Score",
        x = "Factors",
        y = "Correlation with Life Ladder") +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
    theme_minimal(base_size = 8) +
    theme(axis.text.x = element_text(size = 12), 
          axis.text.y = element_text(size = 12), 
          axis.title = element_text(size = 14))

  ggplotly(p)
  ```


  # Let's Draw the Top 3 Happiest and Unhappiest Countries (Total Life Ladder Score) from 2005-2023.
  ```{r}

  total_life_ladder <- world_happiness_clean_main %>%
    group_by(Country.name) %>%
    summarise(Total_Life_Ladder = sum(Life.Ladder, na.rm = TRUE))

  top_3_happiest <- total_life_ladder %>%
    arrange(desc(Total_Life_Ladder)) %>%
    head(3) %>%
    mutate(Category = "Happiest")

  top_3_unhappiest <- total_life_ladder %>%
    arrange(Total_Life_Ladder) %>%
    head(3) %>%
    mutate(Category = "Unhappiest")

  top_countries <- bind_rows(top_3_happiest, top_3_unhappiest)

  p <- ggplot(top_countries, aes(x = reorder(Country.name, Total_Life_Ladder), y = Total_Life_Ladder, fill = Category)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = "Top 3 Happiest and Unhappiest Countries (Total Life Ladder Score)", 
        x = "Country",
        y = "Total Life Ladder Score") +
    scale_fill_manual(values = c("Happiest" = "#208a1a", "Unhappiest" = "#d6b119cd")) +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(size = 12), 
          axis.text.y = element_text(size = 12), 
          axis.title = element_text(size = 14),
          legend.title = element_blank())

  p_interactive <- ggplotly(p)

  p_interactive

  ```



  # Let's see the top 3 countries that have had the biggest increase in Life Ladder score from the first year to the last year and put both of them in one graph
  ```{r}

  # Calculate the change in Life Ladder score for each country
  country_changes <- world_happiness_clean %>%
    group_by(Country.name) %>%
    summarise(
      first_year = min(year),
      last_year = max(year),
      first_year_score = Life.Ladder[year == first_year],
      last_year_score = Life.Ladder[year == last_year],
      score_change = last_year_score - first_year_score
    )

  # Find the top 3 countries with the largest increase in Life Ladder score
  top_3_increase <- country_changes %>%
    arrange(desc(score_change)) %>%
    head(3) %>%
    mutate(Change.Type = "Increase")

  # Find the top 3 countries with the largest decrease in Life Ladder score
  top_3_decrease <- country_changes %>%
    arrange(score_change) %>%
    head(3) %>%
    mutate(Change.Type = "Decrease")

  # Combine the top 3 increases and decreases into one data frame
  top_changes <- bind_rows(top_3_increase, top_3_decrease)

  # Visualize the results with a bar plot
  interactive <- ggplot(top_changes, aes(x = reorder(Country.name, score_change), y = score_change, fill = Change.Type)) +
    geom_bar(stat = "identity") +
    labs(title = "Top 3 Countries with Largest Increase and Decrease in Life Ladder Score",
        x = "Country",
        y = "Change in Life Ladder Score") +
    theme_minimal(base_size = 8) +
    theme(axis.text.x = element_text(size = 12, angle = 90, hjust = 1), 
          axis.text.y = element_text(size = 12), 
          axis.title = element_text(size = 14)) +
    scale_fill_manual(values = c("Increase" = "darkblue", "Decrease" = "orange"))

  ggplotly(interactive)
  ```


  # Let's see the the factor that made the biggest change for for the top country increased (Nicaragua)
  ```{r}
  install.packages("tidyr")
  library(tidyr)

  afghanistan_data <- world_happiness_clean_main %>%
    filter(Country.name == "Afghanistan")

  # Calculate the change in Life Ladder score over time
  score_changes <- afghanistan_data %>%
    arrange(year) %>%
    mutate(score_change = Life.Ladder - lag(Life.Ladder, default = first(Life.Ladder)))

  # Identify the factor with the highest correlation with score_change
  correlations <- cor(score_changes[, c("score_change", "Log.GDP.per.capita", "Social.support", 
                                        "Healthy.life.expectancy.at.birth", "Freedom.to.make.life.choices", 
                                        "Generosity", "Perceptions.of.corruption", "Positive.affect", 
                                        "Negative.affect")], use = "complete.obs")

  # Find the factor with the highest correlation with score_change
  most_impactful_factor <- names(which.max(abs(correlations["score_change", -1])))

  # Print the most impactful factor
  most_impactful_factor


  ```


  # Let's see the the factor that made the biggest change for for the top country Decreased (Afghanistan)
  ```{r}

  # Filter data for Afghanistan
  afghanistan_data <- world_happiness_clean_main %>%
    filter(Country.name == "Afghanistan")

  # Calculate the change in Life Ladder score over time
  afghanistan_data <- afghanistan_data %>%
    arrange(year) %>%
    mutate(score_change = Life.Ladder - lag(Life.Ladder, default = first(Life.Ladder)))

  # Calculate correlations between score_change and other factors
  correlations <- cor(score_changes[, c("score_change", "Log.GDP.per.capita", "Social.support", 
                                        "Healthy.life.expectancy.at.birth", "Freedom.to.make.life.choices", 
                                        "Generosity", "Perceptions.of.corruption", "Positive.affect", 
                                        "Negative.affect")], use = "complete.obs")
  # Find the factor with the highest correlation with score_change
  most_impactful_factor <- names(which.max(abs(correlations["score_change", -1])))

  # Print the most impactful factor
  most_impactful_factor

  ```

# Here is a link to some of my interesting visualizations of the World_Happiness_Report on Tableau Public: (https://public.tableau.com/app/profile/moath.alshannaq/viz/WorldHappinessReportAnalysis2005-2024/LifeLadderScoreDistributionforEachCountryFrom2005-2023)



## Conclusion

# In this analysis, I found the following:

# The three most significant factors affecting country happiness are Log GDP per Capita, Social Support, and Healthy Life Expectancy at Birth.

# Top three happiest countries are Denmark, Sweden, and Canada, and the top three unhappiest countries are Guyana, Cuba, and Maldives.

# The top three countries with the biggest increase in happiness are Bulgaria, Kosovo, and Nicaragua, while the top three with the biggest decrease are Afghanistan, Jordan, and Lebanon.

# For Nicaragua, the factor with the most significant impact on its happiness increase was Social Support.

# For Afghanistan, the factor contributing most to its decrease in happiness was also Social Support.

# I have included links above to Tableau visuals in the notebook, which offer helpful visualizations such as the correlation between happiness rank and various factors, a map showing the distribution of happiness worldwide, and other insightful charts.
